---
title: "rankflex: Quick Tutorial"
author: "Shiyu Xiong, Ruowei Fang, Yang Li"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rankflex: Quick Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  warning  = FALSE,
  message  = FALSE
)
```

## Overview

**rankflex** provides a flexible framework for integrating, aggregating, and
visualizing rankings from multiple sources. The built-in sample dataset
(`rankings_sample`) covers global university rankings from QS, Times Higher
Education (THE), and ARWU across 2023–2025, but every function works with
any ranked domain — films, restaurants, sports teams, and so on.

The typical workflow is:

```
load_rankings()  ->  clean_rankings()  ->  merge_datasets()
      ->  standardized_scores()  ->  compute_composite()
      ->  compare_rankings() / sensitivity_analysis() / optimize_weights()
      ->  plot_ranking_static() / plot_ranking_dynamic()
```

---

## 1. The built-in dataset

`rankings_sample` is a **long-format** data frame: each row is one
university × one source × one year.  
Indicator columns belonging to *other* sources are `NA` for that row
(e.g. `score_teaching` is only filled for THE rows).

```{r load}
library(rankflex)
data(rankings_sample)
cat("Dimensions:", nrow(rankings_sample), "rows ×",
    ncol(rankings_sample), "columns\n")
cat("Sources:", paste(unique(rankings_sample$source), collapse = ", "), "\n")
cat("Years  :", paste(sort(unique(rankings_sample$year)),  collapse = ", "), "\n")
```

```{r glimpse}
head(rankings_sample[rankings_sample$source == "QS" &
                     rankings_sample$year   == 2025,
                     c("entity", "source", "rank",
                       "score_overall", "score_academic_reputation",
                       "score_teaching")], 5)
```

---

## 2. Loading your own data

`load_rankings()` reads a CSV or Excel file and attaches `source` / `year`
metadata columns automatically. `load_data()` is the full version with extra
options (encoding, row-skipping, arbitrary metadata).

```{r load-demo, eval = FALSE}
# Not run — replace with your own file paths
qs_raw   <- load_rankings("qs_2025.csv",            source = "QS",   year = 2025)
the_raw  <- load_rankings("THE_2016_to_2026.csv",   source = "THE",  year = 2025)
arwu_raw <- load_rankings("all_years_combined.csv", source = "ARWU", year = 2025)
```

---

## 3. Cleaning data

`clean_data()` (alias: `clean_rankings()`) auto-detects entity / rank / score
columns, cleans text encoding, converts rank strings like `"=1"` or
`"101–150"` to integers, and drops rows with missing ranks.

```{r clean}
dirty <- data.frame(
  university_name = c(" Harvard University ", " Stanford University ", ""),
  ranking         = c("=1", "2", "3-5"),
  total_score     = c("98.5", "97.2", "90+"),
  stringsAsFactors = FALSE
)
cleaned <- clean_data(dirty, quiet = TRUE)
cleaned
```

When starting from `rankings_sample` (already clean) you can skip this step.

---

## 4. Merging sources into wide format

`merge_datasets()` pivots a list of data frames into a **wide table** with
separate rank/score columns per source (e.g. `rank_QS`, `rank_THE`,
`rank_ARWU`).

> **Required columns:** each data frame in the list must contain
> `entity`, `year`, `source`, `rank`, and **`score`**.  
> `rankings_sample` stores the overall score as `score_overall`, so we must
> create a `score` column first.

```{r merge}
qs_long   <- rankings_sample[rankings_sample$source == "QS",   ]
the_long  <- rankings_sample[rankings_sample$source == "THE",  ]
arwu_long <- rankings_sample[rankings_sample$source == "ARWU", ]

qs_long$score   <- qs_long$score_overall
the_long$score  <- the_long$score_overall
arwu_long$score <- arwu_long$score_overall

final_wide <- merge_datasets(
  df_list = list(qs_long, the_long, arwu_long),
  verbose = FALSE
)

cat("Wide table:", nrow(final_wide), "rows ×", ncol(final_wide), "columns\n")
names(final_wide)
```

```{r merge-preview}
df_2025 <- final_wide[final_wide$year == 2025, ]
complete_2025 <- df_2025[
  !is.na(df_2025$rank_QS) &
  !is.na(df_2025$rank_THE) &
  !is.na(df_2025$rank_ARWU), ]
head(complete_2025[order(complete_2025$rank_QS),
                   c("entity", "rank_QS", "rank_THE", "rank_ARWU")], 10)
```

---

## 5. Standardizing scores & computing a composite

### Why filter to one source first?

`rankings_sample` is long format — the same university appears up to **three
times per year**.  
Each row only has indicator values for *its own* source; cross-source
columns are `NA`. If you run `standardized_scores()` on the full dataset,
most cells in the `_std` columns will be `NA`, and `compute_composite()` will
return empty results.

**Solution:** filter to a single source before standardizing and computing.

```{r standardize}
qs_df <- rankings_sample[rankings_sample$source == "QS", ]
cat("QS rows:", nrow(qs_df), "\n")
qs_indicators <- get_indicators(qs_df)
cat("QS indicators:\n")
cat(paste(" •", qs_indicators), sep = "\n")
```

```{r standardize-apply}
chosen <- c(
  "score_overall",
  "score_academic_reputation",
  "score_employer_reputation",
  "score_faculty_student",
  "score_citations_faculty",
  "score_intl_students"
)
chosen <- chosen[chosen %in% names(qs_df)]
qs_std <- standardized_scores(qs_df, indicators = chosen)
std_cols <- paste0(chosen, "_std")
sapply(qs_std[, std_cols, drop = FALSE],
       function(x) round(range(x, na.rm = TRUE), 2))
```

### Compute composite scores

```{r composite}
weights <- c(
  score_overall                = 0.25,
  score_academic_reputation    = 0.25,
  score_employer_reputation    = 0.15,
  score_faculty_student        = 0.15,
  score_citations_faculty      = 0.15,
  score_intl_students          = 0.05
)
weights <- weights[names(weights) %in% chosen]
composite_df <- compute_composite(qs_std, weights = weights)
top10 <- composite_df[composite_df$year == 2025 &
                      !is.na(composite_df$composite_rank), ]
top10 <- top10[order(top10$composite_rank), ]
head(top10[, c("entity", "rank", "composite_score", "composite_rank")], 10)
```

---

## 6. Comparing composite vs. original rankings

`compare_rankings()` adds a signed difference column and labels each
university as `over_rated`, `under_rated`, or `same` *from the
perspective of the original source*.

```{r compare}
comp_2025 <- composite_df[composite_df$year == 2025 &
                           !is.na(composite_df$composite_rank), ]
compared <- compare_rankings(comp_2025, source_ranks = "rank")

under <- compared[compared$label_rank == "under_rated", ]
under <- under[order(under$diff_rank), ]
head(under[, c("entity", "rank", "composite_rank", "diff_rank")], 8)
```

```{r compare-over}
over <- compared[compared$label_rank == "over_rated", ]
over <- over[order(over$diff_rank, decreasing = TRUE), ]
head(over[, c("entity", "rank", "composite_rank", "diff_rank")], 8)
```

---

## 7. Sensitivity analysis

`sensitivity_analysis()` perturbs the weights randomly (±20% by default)
and reports how much each university's rank fluctuates. Low `sd_rank` means
the ranking is robust across a wide range of weight choices.

```{r sensitivity, message = TRUE}
sa <- sensitivity_analysis(
  data    = qs_std,
  weights = weights,
  n_sim   = 100,
  top_n   = 15,
  year    = 2025
)
sa
```

---

## 8. Optimizing weights for a target university

`optimize_weights()` uses linear programming (**lpSolve**) to find the
weight vector that *maximizes* a given university's composite score,
subject to weights summing to 1 and no single weight exceeding `max_weight`.

```{r optimize, message = TRUE}
# Which indicator set gives Peking University its best possible ranking?
opt <- optimize_weights(
  data       = qs_std,
  target     = "Peking University",
  indicators = chosen,
  year       = 2025,
  max_weight = 0.5,
  top_n      = 20
)
cat("Optimal weights:\n")
print(round(opt$optimal_weights, 4))
cat(sprintf(
  "\nPeking University under optimal weights:\n  Composite score : %.2f\n  Composite rank  : %d\n",
  opt$target_score, opt$target_rank
))
```

```{r optimize-table}
head(opt$ranking[order(opt$ranking$composite_rank),
                 c("entity", "composite_score", "composite_rank")], 10)
```

---

## 9. Static visualization

`plot_ranking_static()` returns a **ggplot2** horizontal bar chart of the
top-ranked universities in a given year.

```{r plot-static, fig.cap = "Top 20 universities by composite score (QS 2025)"}
p <- plot_ranking_static(
  composite_df,
  year  = 2025,
  top_n = 20,
  title = "QS 2025 Top 20 Composite Scores"
)
p
```

---

## 10. Animated ranking

`plot_ranking_dynamic()` creates a racing bar chart showing how rankings
evolve over the three years. It requires **gganimate** and **gifski**.

```{r plot-dynamic, eval = FALSE}
# install.packages(c("gganimate", "gifski"))   # if not yet installed
anim <- plot_ranking_dynamic(composite_df, top_n = 10, fps = 5, duration = 10)
gganimate::animate(anim, fps = 5, duration = 10)
gganimate::anim_save("university_ranking_race.gif", anim)
```

---

## 11. Using multiple sources together

If you want a composite that draws on indicators from **QS, THE, and ARWU
simultaneously**, join the wide table back to the long-format data, or
build a merged indicator matrix manually.

```{r multi-source}
tri_source <- complete_2025[, c("entity", "year",
                                "rank_QS", "rank_THE", "rank_ARWU",
                                "score_QS", "score_THE", "score_ARWU")]

tri_std <- standardized_scores(
  tri_source,
  indicators = c("score_QS", "score_THE", "score_ARWU")
)

tri_weights <- c(score_QS = 1/3, score_THE = 1/3, score_ARWU = 1/3)
tri_composite <- compute_composite(tri_std, weights = tri_weights)

tri_top <- tri_composite[!is.na(tri_composite$composite_rank), ]
tri_top  <- tri_top[order(tri_top$composite_rank), ]
head(tri_top[, c("entity", "rank_QS", "rank_THE", "rank_ARWU",
                 "composite_score", "composite_rank")], 10)
```

---

## Quick-reference: column name rules

| Situation | What to pass |
|-----------|-------------|
| `standardized_scores(df, indicators)` | original column names, e.g. `"score_overall"` |
| `compute_composite(df, weights)` | names of **weight vector** = original indicator names (no `_std`) |
| `merge_datasets(df_list)` | each df must have `entity`, `year`, `source`, `rank`, **`score`** |
| `compare_rankings(df, source_ranks)` | column name(s) of original rank, e.g. `"rank"` |
| `sensitivity_analysis()` / `optimize_weights()` | same `_std`-equipped df as `compute_composite()` |

---

## Package advantages and course takeaways

Developing **rankflex** brought together nearly every major topic covered in
STA3005, and the process crystallised several important lessons about
writing production-quality R code.

**1. The pipe operator `%>%` makes multi-step pipelines readable.**  
Functions like `compute_composite()` use `dplyr`'s `%>%` internally to
chain `group_by()`, `mutate()`, and `ungroup()` into a single, readable
expression. Without the pipe, deeply nested function calls become almost
impossible to follow or debug. Writing the package made clear why `%>%` is
a first-class citizen in modern R.

**2. `ggplot2`'s grammar of graphics enables expressive, reusable charts.**  
`plot_ranking_static()` and `plot_ranking_dynamic()` are built entirely with
`ggplot2` layers — `geom_col()`, `geom_text()`, `coord_flip()`,
`scale_fill_gradient()` — composing them like sentences. This declarative
approach means adding a new visual element is a one-line change rather than
a full rewrite, a stark contrast to base-R graphics.

**3. Roxygen2 keeps code and documentation in sync.**  
Writing `@param`, `@return`, and `@examples` tags directly above each
function means documentation never drifts from the implementation.
`devtools::document()` then auto-generates the NAMESPACE and `.Rd` files,
eliminating a whole class of maintenance bugs that arise when docs live in
separate files.

**4. Separating concerns with internal helpers improves testability.**  
The file-reading logic lives in the unexported `.safe_read_file()`, hidden
from the user. This allowed `load_data()` behaviour — metadata attachment,
error on a missing file — to be tested independently of encoding-detection
complexity, a direct application of the single-responsibility principle.

**5. `testthat` turns assumptions into guarantees.**  
Before adding tests, a subtle bug lurked undetected: passing the full
long-format `rankings_sample` to `compute_composite()` silently produced
0 rows because cross-source `NA` values poisoned the `_std` columns.
A test asserting `nrow(result) > 0` caught this immediately and forced a
cleaner data-flow design (filter to one source before standardising).

**6. Linear programming (`lpSolve`) turns an open question into a precise answer.**  
`optimize_weights()` frames "what weighting scheme makes university X look
best?" as an LP problem — maximise a linear objective subject to simplex
constraints. This demonstrates that R packages need not be limited to
estimation and inference; optimisation belongs here too.

**7. `gganimate` extends the grammar of graphics to time.**  
`plot_ranking_dynamic()` adds just three `gganimate` layers
(`transition_states()`, `ease_aes()`, `enter_fade()`) to an otherwise
standard `ggplot2` object to produce a racing bar chart. Because `gganimate`
is built on top of `ggplot2`'s layer system, the conceptual leap from static
to animated is surprisingly small.

**8. Domain-agnostic design requires deliberate abstraction.**  
Early versions hard-coded university-specific column names everywhere.
Refactoring to accept user-supplied `entity_patterns`, `rank_patterns`, and
`source` strings — and validating with synthetic non-university data —
forced a clean separation between *what the function does* and *what the
data means*, which is the essence of reusable software.

**9. `dplyr` verbs express grouped operations more safely than base R.**  
`compute_composite()` assigns composite ranks *within each year* using
`group_by(year) %>% mutate(composite_rank = min_rank(-composite_score))`.
The equivalent base-R approach with `ave()` or `tapply()` is more verbose
and easier to get wrong when groups contain ties. `dplyr`'s verb-based API
makes intent explicit and handles edge cases consistently.

**10. `LazyData: true` and `usethis::use_data()` make bundled datasets frictionless.**  
Packaging the cleaned `rankings_sample` dataset with
`usethis::use_data(..., compress = "xz")` and declaring `LazyData: true` in
DESCRIPTION gives users instant `data(rankings_sample)` access with zero
loading overhead until the object is actually referenced — a small
ergonomic detail that makes vignettes and examples dramatically easier to
write and run.
